---
title: "Manuscript_draft"
output: html_document
date: "2026-04-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(here)
library(tidyverse)
library(eyemovements)  
library(mltools)
library(dplyr)
library(ggplot2)
library(eyemovements)
library(data.table)
library(patchwork)
library(signal)
library(T4transport)


```

## Introduction
This study evaluates how well traditional and machine-learning-based eye-movement event detection algorithms detect fixations and saccades in noisy, low-sampling-rate webcam eye-tracking data, using human-labelled ground truth as the reference.

## Method

### Dataset

```{r, include= F }

dat <- read_csv(here("data/manual_labels", "webdata_manual_labels.csv"))

n_before_blinks<- nrow(dat)

# Clean data:
dat<- subset(dat,conf>0 & el_pupil>0)# remove blinks:

n_after_blinks<- nrow(dat)

blink_loss<- n_after_blinks/ n_before_blinks

labels<- read_csv(here("data", "manual_ground_truth.csv"))

n_fixations<- length(which(labels$label=="fixation"))
n_saccades<- length(which(labels$label=="saccade"))
n_pso<- length(which(labels$label=="pso"))
n_unclear<- length(which(labels$label=="unclear"))

```

The data contained `r n_before_blinks` raw sample observatations. After blinks were removed, `r n_after_blinks` remained (`r round(blink_loss*100,2)` %). The data manually labelled ground truth contained `r n_fixations` fixations, `r n_saccades` saccades, `r n_pso` PSOs, and `r n_unclear` unclear events (noise). 


```{r}
labels %>% 
  dplyr::filter(!label %in% c("pso", "unclear")) %>%
  ggplot(aes(x = label, y = duration, fill = label)) +
  geom_violin(alpha = 0.4) +
  geom_boxplot(width = 0.1) +
  theme_bw(18)
```


### Saccade Detection algorithms

#### I-VT

Fixation detection

The velocity-threshold identification algorithm (I-VT) was used to detect fixations and saccades from the webcam gaze samples. I-VT classifies eye-movement events based on the point-to-point velocity of gaze samples. In the present analysis, I-VT was implemented using the `IVT()` function from the `eyemovements` package.  A range of velocity thresholds was tested from 20 to 70 deg/s in steps of 5 deg/s. with the minimum fixation duration set to 50 ms. For fixation detection, `IVT()` was run with `return_saccades = FALSE`. The detected fixation intervals were then converted from event-level output to sample-level labels by joining each detected fixation back to the original time series using `fix_start` and `fix_end`. For evaluation, human-labelled fixation samples were coded as 1 and all other samples were coded as 0. The algorithm output was coded in the same binary format, and performance was assessed using the Matthews correlation coefficient (MCC) across all subjects and trials.


```{r}
# IVT-fixation

# Define velocity thresholds to test
vel_thresholds <- seq(20, 70, by = 5)

# Get all subject-trial combinations
trial_list <- dat %>%
  distinct(sub, Trial_Id)

# Loop over each velocity threshold
ivt_fixation_results_mean <- lapply(vel_thresholds, function(v) {

  all_d2 <- list()

  for(i in 1:nrow(trial_list)) {

    # Select one trial
    d <- subset(
      dat,
      sub == trial_list$sub[i] &
        Trial_Id == trial_list$Trial_Id[i]
    )

    # Keep required columns
    d <- d %>%
      transmute(
        x = x,
        y = y,
        time = time_start,
        GT = ground_truth
      )

    # Skip very short trials
    if(nrow(d) < 5) next

    # Smooth gaze samples
    d <- SmoothSamples(d, method = "Mean", window_size = 3)

    # Run IVT fixation detection
    algo_IVT <- IVT(
      data = d,
      vel_threshold = v,
      min_fix_dur = 50,
      return_saccades = FALSE,
      dva_x = 0.01865554,
      dva_y = 0.01919689
    )

    # Convert detected fixations to sample-level labels
    if(nrow(algo_IVT) == 0) {

      d_IVT <- d %>%
        mutate(algo_state = "none") %>%
        select(x, y, time, GT, algo_state)

    } else {

      d_IVT <- d %>%
        left_join(
          algo_IVT %>%
            transmute(
              fix_start,
              fix_end,
              algo_state = "fixation"
            ),
          join_by(
            between(time, fix_start, fix_end)
          )
        ) %>%
        mutate(
          algo_state = ifelse(
            is.na(algo_state),
            "none",
            algo_state
          )
        ) %>%
        select(x, y, time, GT, algo_state)
    }

    # Combine algorithm labels with ground truth
    d2 <- d
    d2$IVT <- d_IVT$algo_state

    # Convert labels to binary fixation classes
    d2 <- d2 %>%
      mutate(
        GT_bin = ifelse(GT == "fixation", 1, 0),
        IVT_bin = ifelse(IVT == "fixation", 1, 0)
      )

    all_d2[[i]] <- d2
  }

  # Combine all trials for this threshold
  all_d2 <- bind_rows(all_d2)

  # Calculate MCC
  data.frame(
    method = "IVT",
    event = "fixation",
    vel_threshold = v,
    MCC = mcc(
      actuals = all_d2$GT_bin,
      preds = all_d2$IVT_bin
    )
  )
})

# Combine results across thresholds
ivt_fixation_results_mean <- bind_rows(ivt_fixation_results_mean)

print(ivt_fixation_results_mean)

# Plot MCC across velocity thresholds
ggplot(ivt_fixation_results_mean, aes(x = vel_threshold, y = MCC)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  labs(
    title = "IVT Fixation MCC Across Velocity Thresholds",
    x = "Velocity threshold",
    y = "Fixation MCC"
  ) +
  theme_classic(base_size = 14)
```
Saccade detection

The same I-VT procedure was also used to detect saccades from the cleaned webcam gaze samples. The cleaned and smoothed gaze coordinates were analysed using the `IVT()` function with the same velocity-threshold range, from 20 to 70 deg/s in steps of 5 deg/s. For saccade detection, the algorithm was run with `return_saccades = TRUE`, while retaining the same minimum fixation duration setting of 50 ms and the package’s default minimum saccade amplitude criterion. Detected saccade intervals were mapped back to sample-level labels using `sacc_start` and `sacc_end`. For evaluation, human-labelled saccade samples were coded as 1 and all other samples were coded as 0. The I-VT output was converted into the same binary format, and MCC was calculated across all subjects and trials.


```{r}
# IVT-saccade

# Loop over each velocity threshold
ivt_saccade_results_mean <- lapply(vel_thresholds, function(v) {

  all_d2 <- list()

  for(i in 1:nrow(trial_list)) {

    # Select one trial
    d <- subset(
      dat,
      sub == trial_list$sub[i] &
        Trial_Id == trial_list$Trial_Id[i]
    )

    # Keep required columns
    d <- d %>%
      transmute(
        x = x,
        y = y,
        time = time_start,
        GT = ground_truth
      )

    # Skip very short trials
    if(nrow(d) < 5) next

    # Smooth gaze samples
    d <- SmoothSamples(d, method = "Mean", window_size = 3)

    # Run IVT saccade detection
    algo_IVT <- IVT(
      data = d,
      vel_threshold = v,
      min_fix_dur = 50,
      return_saccades = TRUE,
      dva_x = 0.01865554,
      dva_y = 0.01919689
    )

    # Convert detected saccades to sample-level labels
    if(nrow(algo_IVT) == 0) {

      d_IVT <- d %>%
        mutate(algo_state = "none") %>%
        select(x, y, time, GT, algo_state)

    } else {

      d_IVT <- d %>%
        left_join(
          algo_IVT %>%
            transmute(
              sacc_start,
              sacc_end,
              algo_state = "saccade"
            ),
          join_by(
            between(time, sacc_start, sacc_end)
          )
        ) %>%
        mutate(
          algo_state = ifelse(
            is.na(algo_state),
            "none",
            algo_state
          )
        ) %>%
        select(x, y, time, GT, algo_state)
    }

    # Combine algorithm labels with ground truth
    d2 <- d
    d2$IVT <- d_IVT$algo_state

    # Convert labels to binary saccade classes
    d2 <- d2 %>%
      mutate(
        GT_bin = ifelse(GT == "saccade", 1, 0),
        IVT_bin = ifelse(IVT == "saccade", 1, 0)
      )

    all_d2[[i]] <- d2
  }

  # Combine all trials for this threshold
  all_d2 <- bind_rows(all_d2)

  # Calculate MCC
  data.frame(
    method = "IVT",
    event = "saccade",
    vel_threshold = v,
    MCC = mcc(
      actuals = all_d2$GT_bin,
      preds = all_d2$IVT_bin
    )
  )
})

# Combine results across thresholds
ivt_saccade_results_mean <- bind_rows(ivt_saccade_results_mean)

print(ivt_saccade_results_mean)

# Plot MCC across velocity thresholds
ggplot(ivt_saccade_results_mean, aes(x = vel_threshold, y = MCC)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  labs(
    title = "IVT Saccade MCC Across Velocity Thresholds",
    x = "Velocity threshold",
    y = "Saccade MCC"
  ) +
  theme_classic(base_size = 14)
```



#### I-DT


The dispersion-threshold identification algorithm (I-DT) was also used to detect eye-movement events. Unlike I-VT, which classifies eye movements based on velocity, I-DT identifies fixations based on the spatial dispersion of gaze samples within a temporal window. The basic assumption is that gaze samples belonging to a fixation should remain spatially close to each other, whereas larger dispersion indicates movement between fixation locations. In the present analysis, I-DT was implemented using the `IDT()` function from the `eyemovements` package and applied to the cleaned webcam gaze samples. Before detection, gaze coordinates were smoothed using a three-sample moving average window. Dispersion thresholds from 1 to 8 degrees were tested in steps of 0.5 degrees, and the temporal window threshold was set to 50 ms. Detected fixation intervals were converted to sample-level labels by mapping each event back to the original time series using `fix_start` and `fix_end`. Human-labelled fixation samples were coded as 1 and all other samples as 0, and the algorithm output was coded in the same way. Performance was evaluated using the Matthews correlation coefficient (MCC) across all subjects and trials.


```{r}
# IDT-fixation

# Define dispersion thresholds to test
disp_thresholds <- seq(1, 8, by = 0.5)

# Get all subject-trial combinations
trial_list <- dat %>%
  distinct(sub, Trial_Id)

# Loop over each dispersion threshold
idt_fixation_results_mean <- lapply(disp_thresholds, function(disp) {
  
  all_d2 <- list()
  
  for(i in 1:nrow(trial_list)) {
    
    # Select one trial
    d <- subset(
      dat,
      sub == trial_list$sub[i] &
        Trial_Id == trial_list$Trial_Id[i]
    )
    
    # Keep required columns
    d <- d %>%
      transmute(
        x = x,
        y = y,
        time = time_start,
        GT = ground_truth
      )
    
    # Skip very short trials
    if(nrow(d) < 5) next
    
    # Smooth gaze samples
    d <- SmoothSamples(d, method = "Mean", window_size = 3)
    
    # Run IDT fixation detection
    algo_IDT <- IDT(
      data = d,
      disp_threshold = disp,
      window_threshold = 50,
      dva_x = 0.01865554,
      dva_y = 0.01919689
    )
    
    # Convert detected fixations to sample-level labels
    if(nrow(algo_IDT) == 0) {
      
      d_IDT <- d %>%
        mutate(algo_state = "none") %>%
        select(x, y, time, GT, algo_state)
      
    } else {
      
      d_IDT <- d %>%
        left_join(
          algo_IDT %>%
            transmute(
              fix_start,
              fix_end,
              algo_state = "fixation"
            ),
          join_by(
            between(time, fix_start, fix_end)
          )
        ) %>%
        mutate(
          algo_state = ifelse(
            is.na(algo_state),
            "none",
            algo_state
          )
        ) %>%
        select(x, y, time, GT, algo_state)
    }
    
    # Combine algorithm labels with ground truth
    d2 <- d
    d2$IDT <- d_IDT$algo_state
    
    # Convert labels to binary fixation classes
    d2 <- d2 %>%
      mutate(
        GT_bin = ifelse(GT == "fixation", 1, 0),
        IDT_bin = ifelse(IDT == "fixation", 1, 0)
      )
    
    all_d2[[i]] <- d2
  }
  
  # Combine all trials for this threshold
  all_d2 <- bind_rows(all_d2)
  
  # Calculate MCC
  data.frame(
    method = "IDT",
    event = "fixation",
    disp_threshold = disp,
    MCC = mcc(
      actuals = all_d2$GT_bin,
      preds = all_d2$IDT_bin
    )
  )
})

# Combine results across thresholds
idt_fixation_results_mean <- bind_rows(idt_fixation_results_mean)

print(idt_fixation_results_mean)

# Plot MCC across dispersion thresholds
ggplot(idt_fixation_results_mean, aes(x = disp_threshold, y = MCC)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  labs(
    title = "IDT Fixation MCC Across Dispersion Thresholds",
    x = "Dispersion threshold",
    y = "Fixation MCC"
  ) +
  theme_classic(base_size = 14)
```
For saccade detection, the same I-DT procedure was applied to the cleaned webcam gaze samples, using the same range of dispersion thresholds from 1 to 8 degrees in steps of 0.5 degrees and a temporal window threshold of 50 ms. The algorithm was run with `return_saccades = TRUE`, and the detected saccade intervals were converted to sample-level labels using `sacc_start` and `sacc_end`. Human-labelled saccade samples were coded as 1 and all other samples as 0, and the algorithm output was coded in the same binary format. Performance was evaluated using MCC across all subjects and trials.


```{r}
# IDT-saccade
idt_saccade_results_mean <- lapply(disp_thresholds, function(disp) {
  
  all_d2 <- list()
  
  for(i in 1:nrow(trial_list)) {
    
    # Select one subject-trial
    d <- subset(
      dat,
      sub == trial_list$sub[i] &
        Trial_Id == trial_list$Trial_Id[i]
    )
    
    # Keep required columns
    d <- d %>%
      transmute(
        x = x,
        y = y,
        time = time_start,
        GT = ground_truth
      )
    
    # Skip very short trials
    if(nrow(d) < 5) next
    
    # Smooth webcam gaze samples
    d <- SmoothSamples(d, method = "Mean", window_size = 3)
    
    # Run IDT and return saccades
    algo_IDT <- tryCatch(
      {
        IDT(
          data = d,
          disp_threshold = disp,
          window_threshold = 50,
          return_saccades = TRUE,
          dva_x = 0.01865554,
          dva_y = 0.01919689
        )
      },
      error = function(e) {
        return(NULL)
      }
    )
    
    # Skip trials where IDT failed
    if(is.null(algo_IDT)) {
      next
    }
    
    # Convert detected saccades to sample-level labels
    if(nrow(algo_IDT) == 0) {
      
      d_IDT <- d %>%
        mutate(algo_state = "none") %>%
        select(x, y, time, GT, algo_state)
      
    } else {
      
      d_IDT <- d %>%
        left_join(
          algo_IDT %>%
            transmute(
              sacc_start,
              sacc_end,
              algo_state = "saccade"
            ),
          join_by(
            between(time, sacc_start, sacc_end)
          )
        ) %>%
        mutate(
          algo_state = ifelse(
            is.na(algo_state),
            "none",
            algo_state
          )
        ) %>%
        select(x, y, time, GT, algo_state)
    }
    
    # Combine algorithm labels with ground truth
    d2 <- d
    d2$IDT <- d_IDT$algo_state
    
    # Convert labels to binary saccade classes
    d2 <- d2 %>%
      mutate(
        GT_bin = ifelse(GT == "saccade", 1, 0),
        IDT_bin = ifelse(IDT == "saccade", 1, 0)
      )
    
    all_d2[[length(all_d2) + 1]] <- d2
  }
  
  # Combine all trials for this parameter value
  all_d2 <- bind_rows(all_d2)
  
  # Calculate MCC
  data.frame(
    method = "IDT",
    event = "saccade",
    disp_threshold = disp,
    MCC = mcc(
      actuals = all_d2$GT_bin,
      preds = all_d2$IDT_bin
    )
  )
})

idt_saccade_results_mean <- bind_rows(idt_saccade_results_mean)

print(idt_saccade_results_mean)

# Plot MCC across dispersion thresholds
ggplot(idt_saccade_results_mean, aes(x = disp_threshold, y = MCC)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  labs(
    title = "IDT Saccade MCC Across Dispersion Thresholds",
    x = "Dispersion threshold",
    y = "Saccade MCC"
  ) +
  theme_classic(base_size = 14)

```



#### Engbert & Kliegl, 2003 [EK03]

This algorithm, originally proposed by Engbert and Kliegl (2003), was also applied to the webcam gaze samples. This method is fundamentally a velocity-based saccade detection algorithm: it identifies saccades using an adaptive velocity threshold, with the threshold controlled by the lambda parameter, and treats the intervals between detected saccades as fixations. This is consistent with the implementation documentation, which describes the method as detecting fixations by first applying the Engbert and Kliegl saccade detection algorithm and then classifying the periods between saccades as fixations. In the present analysis, the method was implemented using the `EngbertKliegl03()` function from the `eyemovements` package. Before detection, gaze coordinates were smoothed using a three-sample moving average window. Lambda values from 2 to 15 were tested, and the minimum fixation duration was set to 50 ms. For fixation detection, the algorithm was run with `return_saccades = FALSE`. Detected fixation intervals were converted to sample-level labels by mapping each event back to the original time series using `fix_start` and `fix_end`. Human-labelled fixation samples were coded as 1 and all other samples as 0, and the algorithm output was coded in the same binary format. Performance was evaluated using the Matthews correlation coefficient (MCC) across all subjects and trials.


```{r}
# EK03-fixation

# Define lambda values to test
lambdas <- 2:15

# Get all subject-trial combinations
trial_list <- dat %>%
  distinct(sub, Trial_Id)

# Loop over each lambda value
ek_lambda_results_mean <- lapply(lambdas, function(l) {

  all_d2 <- list()

  for(i in 1:nrow(trial_list)) {

    # Select one trial
    d <- subset(
      dat,
      sub == trial_list$sub[i] &
        Trial_Id == trial_list$Trial_Id[i]
    )

    # Keep required columns
    d <- d %>%
      transmute(
        x = x,
        y = y,
        time = time_start,
        GT = ground_truth
      )

    # Skip very short trials
    if(nrow(d) < 5) next

    # Smooth gaze samples
    d <- SmoothSamples(d, method = "Mean", window_size = 3)

    # Run EK03 fixation detection
    algo_EK <- EngbertKliegl03(
      data = d,
      lambda = l,
      min_fix_dur = 50,
      return_saccades = FALSE,
      dva_x = 0.01865554,
      dva_y = 0.01919689
    )

    # Convert detected fixations to sample-level labels
    d_EK <- d %>%
      left_join(
        algo_EK %>%
          transmute(
            fix_start,
            fix_end,
            algo_state = "fixation"
          ),
        join_by(
          between(time, fix_start, fix_end)
        )
      ) %>%
      mutate(
        algo_state = ifelse(
          is.na(algo_state),
          "none",
          algo_state
        )
      ) %>%
      select(x, y, time, GT, algo_state)

    # Combine algorithm labels with ground truth
    d2 <- d
    d2$EK <- d_EK$algo_state

    # Convert labels to binary fixation classes
    d2 <- d2 %>%
      mutate(
        GT_bin = ifelse(GT == "fixation", 1, 0),
        EK_bin = ifelse(EK == "fixation", 1, 0)
      )

    all_d2[[i]] <- d2
  }

  # Combine all trials for this lambda value
  all_d2 <- bind_rows(all_d2)

  # Calculate MCC
  data.frame(
    lambda = l,
    EK_MCC = mcc(
      actuals = all_d2$GT_bin,
      preds = all_d2$EK_bin
    )
  )
})

# Combine results across lambda values
ek_lambda_results_mean<- bind_rows(ek_lambda_results_mean)

print(ek_lambda_results_mean )

# Plot MCC across lambda values
ggplot(ek_lambda_results_mean, aes(x = lambda, y = EK_MCC)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  labs(
    title = "EK03 Fixation MCC Across Lambda Values",
    x = "Lambda",
    y = "MCC"
  ) +
  theme_classic(base_size = 14)
```

For saccade detection, the same Engbert and Kliegl procedure was applied to the cleaned webcam gaze samples, using lambda values from 2 to 15 and a minimum fixation duration of 50 ms. The algorithm was run with `return_saccades = TRUE`, and detected saccade intervals were converted to sample-level labels using `sacc_start` and `sacc_end`. Human-labelled saccade samples were coded as 1 and all other samples as 0, and the algorithm output was coded in the same binary format. Performance was evaluated using the Matthews correlation coefficient (MCC) across all subjects and trials.

```{r}
# EK03-saccade

# Define lambda values to test
lambdas <- 2:15

# Get all subject-trial combinations
trial_list <- dat %>%
  distinct(sub, Trial_Id)

# Loop over each lambda value
ek_lambda_saccade_results_mean <- lapply(lambdas, function(l) {

  all_d2 <- list()

  for(i in 1:nrow(trial_list)) {

    # Select one trial
    d <- subset(
      dat,
      sub == trial_list$sub[i] &
        Trial_Id == trial_list$Trial_Id[i]
    )

    # Keep required columns
    d <- d %>%
      transmute(
        x = x,
        y = y,
        time = time_start,
        GT = ground_truth
      )

    # Skip very short trials
    if(nrow(d) < 5) next

    # Smooth gaze samples
    d <- SmoothSamples(d, method = "Mean", window_size = 3)

    # Run EK03 saccade detection
    algo_EK <- EngbertKliegl03(
      data = d,
      lambda = l,
      min_fix_dur = 50,
      return_saccades = TRUE,
      dva_x = 0.01865554,
      dva_y = 0.01919689
    )

    # Convert detected saccades to sample-level labels
    if(nrow(algo_EK) == 0) {

      d_EK <- d %>%
        mutate(algo_state = "none") %>%
        select(x, y, time, GT, algo_state)

    } else {

      d_EK <- d %>%
        left_join(
          algo_EK %>%
            transmute(
              sacc_start,
              sacc_end,
              algo_state = "saccade"
            ),
          join_by(
            between(time, sacc_start, sacc_end)
          )
        ) %>%
        mutate(
          algo_state = ifelse(
            is.na(algo_state),
            "none",
            algo_state
          )
        ) %>%
        select(x, y, time, GT, algo_state)
    }

    # Combine algorithm labels with ground truth
    d2 <- d
    d2$EK <- d_EK$algo_state

    # Convert labels to binary saccade classes
    d2 <- d2 %>%
      mutate(
        GT_bin = ifelse(GT == "saccade", 1, 0),
        EK_bin = ifelse(EK == "saccade", 1, 0)
      )

    all_d2[[i]] <- d2
  }

  # Combine all trials for this lambda value
  all_d2 <- bind_rows(all_d2)

  # Calculate MCC
  data.frame(
    lambda = l,
    EK_saccade_MCC = mcc(
      actuals = all_d2$GT_bin,
      preds = all_d2$EK_bin
    )
  )
})

# Combine results across lambda values
ek_lambda_saccade_results_mean <- bind_rows(ek_lambda_saccade_results_mean)

print(ek_lambda_saccade_results_mean)

# Plot MCC across lambda values
ggplot(ek_lambda_saccade_results_mean, aes(x = lambda, y = EK_saccade_MCC)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  labs(
    title = "EK03 Saccade MCC Across Lambda Values",
    x = "Lambda",
    y = "Saccade MCC"
  ) +
  theme_classic(base_size = 14)
```

#### IHMM

The iterative hidden Markov model (IHMM) was also applied to the cleaned webcam gaze samples for fixation detection. The method was implemented using the `IHMM()` function from the `eyemovements` package. As in the other analyses, gaze coordinates were first smoothed using a three-sample moving average window. The algorithm was applied separately to each subject and trial, with the minimum fixation duration set to 50 ms. For fixation detection, IHMM() was run with `return_saccades = FALSE`. The fixation intervals returned by the algorithm were then converted to sample-level labels by mapping each detected event back to the original time series using `fix_start` and `fix_end`. Human-labelled fixation samples were coded as 1 and all other samples as 0, and the IHMM output was coded in the same binary format. Performance was evaluated using the Matthews correlation coefficient (MCC) across all subjects and trials.


```{r}
# IHMM-fixation

# Get all subject-trial combinations
trial_list <- dat %>%
  distinct(sub, Trial_Id)

# Run IHMM across all trials
ihmm_fixation_results_mean <- lapply(1, function(x) {
  
  all_d2 <- list()
  
  for(i in 1:nrow(trial_list)) {
    
    # Select one trial
    d <- subset(
      dat,
      sub == trial_list$sub[i] &
        Trial_Id == trial_list$Trial_Id[i]
    )
    
    # Keep required columns
    d <- d %>%
      transmute(
        x = x,
        y = y,
        time = time_start,
        GT = ground_truth
      )
    
    # Skip very short trials
    if(nrow(d) < 5) next
    
    # Smooth gaze samples
    d <- SmoothSamples(d, method = "Mean", window_size = 3)
    
    # Run IHMM fixation detection
    algo_IHMM <- IHMM(
      data = d,
      min_fix_dur = 50,
      return_saccades = FALSE,
      dva_x = 0.01865554,
      dva_y = 0.01919689
    )
    
    # Convert detected fixations to sample-level labels
    if(nrow(algo_IHMM) == 0) {
      
      d_IHMM <- d %>%
        mutate(algo_state = "none") %>%
        select(x, y, time, GT, algo_state)
      
    } else {
      
      d_IHMM <- d %>%
        left_join(
          algo_IHMM %>%
            transmute(
              fix_start,
              fix_end,
              algo_state = "fixation"
            ),
          join_by(
            between(time, fix_start, fix_end)
          )
        ) %>%
        mutate(
          algo_state = ifelse(
            is.na(algo_state),
            "none",
            algo_state
          )
        ) %>%
        select(x, y, time, GT, algo_state)
    }
    
    # Combine algorithm labels with ground truth
    d2 <- d
    d2$IHMM <- d_IHMM$algo_state
    
    # Convert labels to binary fixation classes
    d2 <- d2 %>%
      mutate(
        GT_bin = ifelse(GT == "fixation", 1, 0),
        IHMM_bin = ifelse(IHMM == "fixation", 1, 0)
      )
    
    all_d2[[i]] <- d2
  }
  
  # Combine all trials
  all_d2 <- bind_rows(all_d2)
  
  # Calculate MCC
  data.frame(
    method = "IHMM",
    event = "fixation",
    MCC = mcc(
      actuals = all_d2$GT_bin,
      preds = all_d2$IHMM_bin
    )
  )
})

# Combine results
ihmm_fixation_results_mean <- bind_rows(ihmm_fixation_results_mean)

print(ihmm_fixation_results_mean)
```
For saccade detection, the same IHMM procedure was applied to the webcam gaze samples. The algorithm was run with `return_saccades = TRUE`, with the minimum fixation duration set to 50 ms. Detected saccade intervals were converted to sample-level labels using `sacc_start` and `sacc_end`. Human-labelled saccade samples were coded as 1 and all other samples as 0, and the IHMM output was coded in the same binary format. Performance was evaluated using the Matthews correlation coefficient (MCC) across all subjects and trials.


```{r}
# IHMM-saccade

# Run IHMM across all trials
ihmm_saccade_results_mean <- lapply(1, function(x) {
  
  all_d2 <- list()
  
  for(i in 1:nrow(trial_list)) {
    
    # Select one trial
    d <- subset(
      dat,
      sub == trial_list$sub[i] &
        Trial_Id == trial_list$Trial_Id[i]
    )
    
    # Keep required columns
    d <- d %>%
      transmute(
        x = x,
        y = y,
        time = time_start,
        GT = ground_truth
      )
    
    # Skip very short trials
    if(nrow(d) < 5) next
    
    # Smooth gaze samples
    d <- SmoothSamples(d, method = "Mean", window_size = 3)
    
    # Run IHMM saccade detection
    algo_IHMM <- IHMM(
      data = d,
      min_fix_dur = 50,
      return_saccades = TRUE,
      dva_x = 0.01865554,
      dva_y = 0.01919689
    )
    
    # Convert detected saccades to sample-level labels
    if(nrow(algo_IHMM) == 0) {
      
      d_IHMM <- d %>%
        mutate(algo_state = "none") %>%
        select(x, y, time, GT, algo_state)
      
    } else {
      
      d_IHMM <- d %>%
        left_join(
          algo_IHMM %>%
            transmute(
              sacc_start,
              sacc_end,
              algo_state = "saccade"
            ),
          join_by(
            between(time, sacc_start, sacc_end)
          )
        ) %>%
        mutate(
          algo_state = ifelse(
            is.na(algo_state),
            "none",
            algo_state
          )
        ) %>%
        select(x, y, time, GT, algo_state)
    }
    
    # Combine algorithm labels with ground truth
    d2 <- d
    d2$IHMM <- d_IHMM$algo_state
    
    # Convert labels to binary saccade classes
    d2 <- d2 %>%
      mutate(
        GT_bin = ifelse(GT == "saccade", 1, 0),
        IHMM_bin = ifelse(IHMM == "saccade", 1, 0)
      )
    
    all_d2[[i]] <- d2
  }
  
  # Combine all trials
  all_d2 <- bind_rows(all_d2)
  
  # Calculate MCC
  data.frame(
    method = "IHMM",
    event = "saccade",
    MCC = mcc(
      actuals = all_d2$GT_bin,
      preds = all_d2$IHMM_bin
    )
  )
})

# Combine results
ihmm_saccade_results_mean <- bind_rows(ihmm_saccade_results_mean)

print(ihmm_saccade_results_mean)
```

### Sample Smoothing methods

Raw, Mean filter, Median filter, Savitsky-Golay, Others? 


Four versions of the webcam gaze signal were compared to examine how different smoothing methods affected event detection performance. The raw version used the original webcam x- and y-coordinates without additional smoothing, providing a baseline for comparison. The moving-average smoothing method applied a three-sample mean window to the gaze coordinates. This method reduces short-term sample-to-sample noise by replacing each gaze sample with the local average of neighbouring samples. Because the webcam data were sampled at 60 Hz, a three-sample window corresponds to a short temporal window and avoids excessive temporal smoothing. The median smoothing method also used a three-sample window, but each value was replaced by the local median rather than the local mean. This makes the median filter less sensitive to brief outlier samples or sudden spikes in the webcam signal. Finally, Savitzky–Golay smoothing was applied using a fifth-order polynomial and a 23-sample window. This method smooths the gaze trajectory while attempting to preserve the overall shape of the signal, such as local rises and falls in position.

For both IVT and IDT, each smoothing method was applied separately to the webcam gaze samples before event detection. The resulting smoothed or unsmoothed coordinates were then used as input to the corresponding algorithm. IVT was tested across velocity thresholds from 20 to 70 deg/s, while IDT was tested across dispersion thresholds from 1 to 8 degrees. For IDT, the temporal window threshold was fixed at 50 ms. For both algorithms, fixation and saccade detection were evaluated separately. Detected event intervals were converted back to sample-level labels and compared with the human-labelled ground truth using the Matthews correlation coefficient (MCC).

```{r}

# Setup 

# Smoothing methods
smooth_methods <- c("raw", "mean", "median", "sg")

# IVT parameters
vel_thresholds <- seq(20, 70, by = 5)
min_fix_dur <- 50

# IDT parameters
disp_thresholds <- seq(1, 8, by = 0.5)
window_threshold <- 50

# Get all subject-trial combinations
trial_list <- dat %>%
  distinct(sub, Trial_Id)

# Select x/y coordinates and apply smoothing temporarily
get_xy <- function(data, smooth){
  
  d <- data %>%
    dplyr::transmute(
      x = x,
      y = y,
      time = time_start,
      GT = ground_truth
    )
  
  if(nrow(d) < 5) {
    return(d)
  }
  
  if(smooth == "raw") {
    
    return(d)
    
  } else if(smooth == "mean") {
    
    d <- eyemovements::SmoothSamples(
      d,
      method = "Mean",
      window_size = 3
    )
    
  } else if(smooth == "median") {
    
    d <- eyemovements::SmoothSamples(
      d,
      method = "Median",
      window_size = 3
    )
    
  } else if(smooth == "sg") {
    
    if(nrow(d) >= 23) {
      d$x <- signal::sgolayfilt(d$x, p = 5, n = 23)
      d$y <- signal::sgolayfilt(d$y, p = 5, n = 23)
    }
  }
  
  return(d)
}

```


Extract ground-truth fixation events
```{r}

# Extract ground-truth fixation events

gt_fixation_events <- dat %>%
  dplyr::transmute(
    sub = sub,
    Trial_Id = Trial_Id,
    time = time_start,
    GT = ground_truth
  ) %>%
  dplyr::arrange(sub, Trial_Id, time) %>%
  dplyr::group_by(sub, Trial_Id) %>%
  dplyr::mutate(
    is_fix = GT == "fixation",
    event_id = cumsum(is_fix != dplyr::lag(is_fix, default = first(is_fix)))
  ) %>%
  dplyr::ungroup() %>%
  dplyr::filter(is_fix) %>%
  dplyr::group_by(sub, Trial_Id, event_id) %>%
  dplyr::summarise(
    fix_start = min(time, na.rm = TRUE),
    fix_end = max(time, na.rm = TRUE),
    duration = fix_end - fix_start,
    .groups = "drop"
  )

gt_fixation_summary <- gt_fixation_events %>%
  dplyr::group_by(sub, Trial_Id) %>%
  dplyr::summarise(
    GT_n_fixations = dplyr::n(),
    GT_mean_fix_dur = mean(duration, na.rm = TRUE),
    GT_sd_fix_dur = sd(duration, na.rm = TRUE),
    .groups = "drop"
  )
```


Extract ground-truth saccade events
```{r}
# Extract ground-truth saccade events

gt_saccade_events <- dat %>%
  dplyr::transmute(
    sub = sub,
    Trial_Id = Trial_Id,
    time = time_start,
    GT = ground_truth
  ) %>%
  dplyr::arrange(sub, Trial_Id, time) %>%
  dplyr::group_by(sub, Trial_Id) %>%
  dplyr::mutate(
    is_sacc = GT == "saccade",
    event_id = cumsum(is_sacc != dplyr::lag(is_sacc, default = first(is_sacc)))
  ) %>%
  dplyr::ungroup() %>%
  dplyr::filter(is_sacc) %>%
  dplyr::group_by(sub, Trial_Id, event_id) %>%
  dplyr::summarise(
    sacc_start = min(time, na.rm = TRUE),
    sacc_end = max(time, na.rm = TRUE),
    duration = sacc_end - sacc_start,
    .groups = "drop"
  )

gt_saccade_summary <- gt_saccade_events %>%
  dplyr::group_by(sub, Trial_Id) %>%
  dplyr::summarise(
    GT_n_saccades = dplyr::n(),
    GT_mean_sacc_dur = mean(duration, na.rm = TRUE),
    GT_sd_sacc_dur = sd(duration, na.rm = TRUE),
    .groups = "drop"
  )
```


IVT-Fixation
```{r}

# IVT-fixation

ivt_fixation_results <- lapply(smooth_methods, function(s) {
  
  lapply(vel_thresholds, function(v) {
    
    cat("Running IVT fixation:", s, v, min_fix_dur, "\n")
    
    all_d2 <- list()
    
    for(i in 1:nrow(trial_list)) {
      
      # Select one trial
      df0 <- subset(
        dat,
        sub == trial_list$sub[i] &
          Trial_Id == trial_list$Trial_Id[i]
      )
      
      # Apply selected smoothing method
      d <- get_xy(df0, s)
      
      # Skip very short trials
      if(nrow(d) < 5) next
      
      # Run IVT fixation detection
      algo_IVT <- tryCatch(
        {
          eyemovements::IVT(
            data = d,
            vel_threshold = v,
            min_fix_dur = min_fix_dur,
            return_saccades = FALSE,
            dva_x = 0.01865554,
            dva_y = 0.01919689
          )
        },
        error = function(e) {
          message("IVT fixation error: ", e$message)
          return(NULL)
        }
      )
      
      # Convert detected fixations to sample-level labels
      if(is.null(algo_IVT) || nrow(algo_IVT) == 0) {
        
        d_IVT <- d %>%
          mutate(algo_state = "none")
        
      } else {
        
        d_IVT <- d %>%
          left_join(
            algo_IVT %>%
              transmute(
                fix_start,
                fix_end,
                algo_state = "fixation"
              ),
            join_by(
              between(time, fix_start, fix_end)
            )
          ) %>%
          mutate(
            algo_state = ifelse(
              is.na(algo_state),
              "none",
              algo_state
            )
          )
      }
      
      # Combine algorithm labels with ground truth
      d2 <- d
      d2$IVT <- d_IVT$algo_state
      
      # Convert labels to binary fixation classes
      d2 <- d2 %>%
        mutate(
          GT_bin = ifelse(GT == "fixation", 1, 0),
          IVT_bin = ifelse(IVT == "fixation", 1, 0)
        )
      
      all_d2[[length(all_d2) + 1]] <- d2
    }
    
    # Combine all trials
    all_d2 <- bind_rows(all_d2)
    
    # Calculate MCC
    data.frame(
      method = "IVT",
      event = "fixation",
      smoothing = s,
      vel_threshold = v,
      min_fix_dur = min_fix_dur,
      MCC = mltools::mcc(
        actuals = all_d2$GT_bin,
        preds = all_d2$IVT_bin
      )
    )
  })
})

ivt_fixation_results <- bind_rows(unlist(ivt_fixation_results, recursive = FALSE))

print(ivt_fixation_results)

ggplot(
  ivt_fixation_results,
  aes(x = vel_threshold, y = MCC, colour = smoothing)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(
    title = "IVT Fixation MCC Across Velocity Thresholds",
    x = "Velocity threshold",
    y = "Fixation MCC",
    colour = "Smoothing"
  ) +
  theme_classic(base_size = 14)
```
IVT-Saccade
```{r}
# IVT-saccade

ivt_saccade_results <- lapply(smooth_methods, function(s) {
  
  lapply(vel_thresholds, function(v) {
    
    cat("Running IVT saccade:", s, v, min_fix_dur, "\n")
    
    all_d2 <- list()
    
    for(i in 1:nrow(trial_list)) {
      
      # Select one trial
      df0 <- subset(
        dat,
        sub == trial_list$sub[i] &
          Trial_Id == trial_list$Trial_Id[i]
      )
      
      # Apply selected smoothing method
      d <- get_xy(df0, s)
      
      # Skip very short trials
      if(nrow(d) < 5) next
      
      # Run IVT saccade detection
      algo_IVT <- tryCatch(
        {
          eyemovements::IVT(
            data = d,
            vel_threshold = v,
            min_fix_dur = min_fix_dur,
            return_saccades = TRUE,
            dva_x = 0.01865554,
            dva_y = 0.01919689
          )
        },
        error = function(e) {
          message("IVT saccade error: ", e$message)
          return(NULL)
        }
      )
      
      # Convert detected saccades to sample-level labels
      if(is.null(algo_IVT) || nrow(algo_IVT) == 0) {
        
        d_IVT <- d %>%
          mutate(algo_state = "none")
        
      } else {
        
        d_IVT <- d %>%
          left_join(
            algo_IVT %>%
              transmute(
                sacc_start,
                sacc_end,
                algo_state = "saccade"
              ),
            join_by(
              between(time, sacc_start, sacc_end)
            )
          ) %>%
          mutate(
            algo_state = ifelse(
              is.na(algo_state),
              "none",
              algo_state
            )
          )
      }
      
      # Combine algorithm labels with ground truth
      d2 <- d
      d2$IVT <- d_IVT$algo_state
      
      # Convert labels to binary saccade classes
      d2 <- d2 %>%
        mutate(
          GT_bin = ifelse(GT == "saccade", 1, 0),
          IVT_bin = ifelse(IVT == "saccade", 1, 0)
        )
      
      all_d2[[length(all_d2) + 1]] <- d2
    }
    
    # Combine all trials
    all_d2 <- bind_rows(all_d2)
    
    # Calculate MCC
    data.frame(
      method = "IVT",
      event = "saccade",
      smoothing = s,
      vel_threshold = v,
      min_fix_dur = min_fix_dur,
      MCC = mltools::mcc(
        actuals = all_d2$GT_bin,
        preds = all_d2$IVT_bin
      )
    )
  })
})

ivt_saccade_results <- bind_rows(unlist(ivt_saccade_results, recursive = FALSE))

print(ivt_saccade_results)

ggplot(
  ivt_saccade_results,
  aes(x = vel_threshold, y = MCC, colour = smoothing)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(
    title = "IVT Saccade MCC Across Velocity Thresholds",
    x = "Velocity threshold",
    y = "Saccade MCC",
    colour = "Smoothing"
  ) +
  theme_classic(base_size = 14)
```
IDT-Fixation
```{r}
# IDT-fixation
idt_fixation_results <- lapply(smooth_methods, function(s) {
  
  lapply(disp_thresholds, function(disp) {
    
    cat("Running IDT fixation:", s, disp, window_threshold, "\n")
    
    all_d2 <- list()
    
    for(i in 1:nrow(trial_list)) {
      
      # Select one trial
      df0 <- subset(
        dat,
        sub == trial_list$sub[i] &
          Trial_Id == trial_list$Trial_Id[i]
      )
      
      # Apply selected smoothing method
      d <- get_xy(df0, s)
      
      # Skip very short trials
      if(nrow(d) < 5) next
      
      # Run IDT fixation detection
      algo_IDT <- tryCatch(
        {
          eyemovements::IDT(
            data = d,
            disp_threshold = disp,
            window_threshold = window_threshold,
            return_saccades = FALSE,
            dva_x = 0.01865554,
            dva_y = 0.01919689
          )
        },
        error = function(e) {
          message("IDT fixation error: ", e$message)
          return(NULL)
        }
      )
      
      # Convert detected fixations to sample-level labels
      if(is.null(algo_IDT) || nrow(algo_IDT) == 0) {
        
        d_IDT <- d %>%
          mutate(algo_state = "none")
        
      } else {
        
        d_IDT <- d %>%
          left_join(
            algo_IDT %>%
              transmute(
                fix_start,
                fix_end,
                algo_state = "fixation"
              ),
            join_by(
              between(time, fix_start, fix_end)
            )
          ) %>%
          mutate(
            algo_state = ifelse(
              is.na(algo_state),
              "none",
              algo_state
            )
          )
      }
      
      # Combine algorithm labels with ground truth
      d2 <- d
      d2$IDT <- d_IDT$algo_state
      
      # Convert labels to binary fixation classes
      d2 <- d2 %>%
        mutate(
          GT_bin = ifelse(GT == "fixation", 1, 0),
          IDT_bin = ifelse(IDT == "fixation", 1, 0)
        )
      
      all_d2[[length(all_d2) + 1]] <- d2
    }
    
    # Combine all trials
    all_d2 <- bind_rows(all_d2)
    
    # Calculate MCC
    data.frame(
      method = "IDT",
      event = "fixation",
      smoothing = s,
      disp_threshold = disp,
      window_threshold = window_threshold,
      MCC = mltools::mcc(
        actuals = all_d2$GT_bin,
        preds = all_d2$IDT_bin
      )
    )
  })
})

idt_fixation_results <- bind_rows(unlist(idt_fixation_results, recursive = FALSE))

print(idt_fixation_results)

ggplot(
  idt_fixation_results,
  aes(x = disp_threshold, y = MCC, colour = smoothing)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(
    title = "IDT Fixation MCC Across Dispersion Thresholds",
    x = "Dispersion threshold",
    y = "Fixation MCC",
    colour = "Smoothing"
  ) +
  theme_classic(base_size = 14)
```

IDT-Saccade
```{r}

# IDT-saccade
idt_saccade_results <- lapply(smooth_methods, function(s) {
  
  lapply(disp_thresholds, function(disp) {
    
    cat("Running IDT saccade:", s, disp, window_threshold, "\n")
    
    all_d2 <- list()
    
    for(i in 1:nrow(trial_list)) {
      
      # Select one trial
      df0 <- subset(
        dat,
        sub == trial_list$sub[i] &
          Trial_Id == trial_list$Trial_Id[i]
      )
      
      # Apply selected smoothing method
      d <- get_xy(df0, s)
      
      # Skip very short trials
      if(nrow(d) < 5) next
      
      # Run IDT saccade detection
      algo_IDT <- tryCatch(
        {
          eyemovements::IDT(
            data = d,
            disp_threshold = disp,
            window_threshold = window_threshold,
            return_saccades = TRUE,
            dva_x = 0.01865554,
            dva_y = 0.01919689
          )
        },
        error = function(e) {
          message("IDT saccade error: ", e$message)
          return(NULL)
        }
      )
      
      # Convert detected saccades to sample-level labels
      if(is.null(algo_IDT) || nrow(algo_IDT) == 0) {
        
        d_IDT <- d %>%
          mutate(algo_state = "none")
        
      } else {
        
        d_IDT <- d %>%
          left_join(
            algo_IDT %>%
              transmute(
                sacc_start,
                sacc_end,
                algo_state = "saccade"
              ),
            join_by(
              between(time, sacc_start, sacc_end)
            )
          ) %>%
          mutate(
            algo_state = ifelse(
              is.na(algo_state),
              "none",
              algo_state
            )
          )
      }
      
      # Combine algorithm labels with ground truth
      d2 <- d
      d2$IDT <- d_IDT$algo_state
      
      # Convert labels to binary saccade classes
      d2 <- d2 %>%
        mutate(
          GT_bin = ifelse(GT == "saccade", 1, 0),
          IDT_bin = ifelse(IDT == "saccade", 1, 0)
        )
      
      all_d2[[length(all_d2) + 1]] <- d2
    }
    
    # Combine all trials
    all_d2 <- bind_rows(all_d2)
    
    # Calculate MCC
    data.frame(
      method = "IDT",
      event = "saccade",
      smoothing = s,
      disp_threshold = disp,
      window_threshold = window_threshold,
      MCC = mltools::mcc(
        actuals = all_d2$GT_bin,
        preds = all_d2$IDT_bin
      )
    )
  })
})

idt_saccade_results <- bind_rows(unlist(idt_saccade_results, recursive = FALSE))

print(idt_saccade_results)

ggplot(
  idt_saccade_results,
  aes(x = disp_threshold, y = MCC, colour = smoothing)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(
    title = "IDT Saccade MCC Across Dispersion Thresholds",
    x = "Dispersion threshold",
    y = "Saccade MCC",
    colour = "Smoothing"
  ) +
  theme_classic(base_size = 14)
 
```

EK03-fixation with different smoothing methods
```{r}
# EK03-fixation

# Define lambda values
lambdas <- 2:15

# Run EK03 fixation detection
ek_fixation_results <- lapply(smooth_methods, function(s) {
  
  lapply(lambdas, function(l) {
    
    cat("Running EK03 fixation:", s, l, "\n")
    
    all_d2 <- list()
    
    for(i in 1:nrow(trial_list)) {
      
      # Select one trial
      df0 <- subset(
        dat,
        sub == trial_list$sub[i] &
          Trial_Id == trial_list$Trial_Id[i]
      )
      
      # Apply selected smoothing method
      d <- get_xy(df0, s)
      
      # Skip very short trials
      if(nrow(d) < 5) next
      
      # Run EK03 fixation detection
      algo_EK <- tryCatch(
        {
          eyemovements::EngbertKliegl03(
            data = d,
            lambda = l,
            min_fix_dur = 50,
            return_saccades = FALSE,
            dva_x = 0.01865554,
            dva_y = 0.01919689
          )
        },
        error = function(e) {
          message("EK03 fixation error: ", e$message)
          return(NULL)
        }
      )
      
      # Convert detected fixations to sample-level labels
      if(is.null(algo_EK) || nrow(algo_EK) == 0) {
        
        d_EK <- d %>%
          dplyr::mutate(algo_state = "none")
        
      } else {
        
        d_EK <- d %>%
          dplyr::left_join(
            algo_EK %>%
              dplyr::transmute(
                fix_start,
                fix_end,
                algo_state = "fixation"
              ),
            dplyr::join_by(
              between(time, fix_start, fix_end)
            )
          ) %>%
          dplyr::mutate(
            algo_state = ifelse(
              is.na(algo_state),
              "none",
              algo_state
            )
          )
      }
      
      # Combine algorithm labels with ground truth
      d2 <- d
      d2$EK <- d_EK$algo_state
      
      # Convert labels to binary fixation classes
      d2 <- d2 %>%
        dplyr::mutate(
          GT_bin = ifelse(GT == "fixation", 1, 0),
          EK_bin = ifelse(EK == "fixation", 1, 0)
        )
      
      all_d2[[length(all_d2) + 1]] <- d2
    }
    
    # Combine all trials
    all_d2 <- dplyr::bind_rows(all_d2)
    
    # Calculate MCC
    data.frame(
      method = "EK03",
      event = "fixation",
      smoothing = s,
      lambda = l,
      min_fix_dur = 50,
      MCC = mltools::mcc(
        actuals = all_d2$GT_bin,
        preds = all_d2$EK_bin
      )
    )
  })
})

# Combine results
ek_fixation_results <- dplyr::bind_rows(
  unlist(ek_fixation_results, recursive = FALSE)
)

print(ek_fixation_results)

# Plot fixation MCC
ggplot(
  ek_fixation_results,
  aes(x = lambda, y = MCC, colour = smoothing)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(
    title = "EK03 Fixation MCC Across Lambda Values",
    x = "Lambda",
    y = "Fixation MCC",
    colour = "Smoothing"
  ) +
  theme_classic(base_size = 14)
```
EK03-saccade
```{r}

# Run EK03 saccade detection
lambdas <- 2:15

ek_saccade_results <- lapply(smooth_methods, function(s) {
  
  lapply(lambdas, function(l) {
    
    cat("Running EK03 saccade:", s, l, "\n")
    
    all_d2 <- list()
    
    for(i in 1:nrow(trial_list)) {
      
      # Select one trial
      df0 <- subset(
        dat,
        sub == trial_list$sub[i] &
          Trial_Id == trial_list$Trial_Id[i]
      )
      
      # Apply selected smoothing method
      d <- get_xy(df0, s)
      
      # Skip very short trials
      if(nrow(d) < 5) next
      
      # Run EK03 saccade detection
      algo_EK <- tryCatch(
        {
          eyemovements::EngbertKliegl03(
            data = d,
            lambda = l,
            min_fix_dur = 50,
            return_saccades = TRUE,
            dva_x = 0.01865554,
            dva_y = 0.01919689
          )
        },
        error = function(e) {
          message("EK03 saccade error: ", e$message)
          return(NULL)
        }
      )
      
      # Convert detected saccades to sample-level labels
      if(is.null(algo_EK) || nrow(algo_EK) == 0) {
        
        d_EK <- d %>%
          dplyr::mutate(algo_state = "none")
        
      } else {
        
        d_EK <- d %>%
          dplyr::left_join(
            algo_EK %>%
              dplyr::transmute(
                sacc_start,
                sacc_end,
                algo_state = "saccade"
              ),
            dplyr::join_by(
              between(time, sacc_start, sacc_end)
            )
          ) %>%
          dplyr::mutate(
            algo_state = ifelse(
              is.na(algo_state),
              "none",
              algo_state
            )
          )
      }
      
      # Combine algorithm labels with ground truth
      d2 <- d
      d2$EK <- d_EK$algo_state
      
      # Convert labels to binary saccade classes
      d2 <- d2 %>%
        dplyr::mutate(
          GT_bin = ifelse(GT == "saccade", 1, 0),
          EK_bin = ifelse(EK == "saccade", 1, 0)
        )
      
      all_d2[[length(all_d2) + 1]] <- d2
    }
    
    # Combine all trials
    all_d2 <- dplyr::bind_rows(all_d2)
    
    # Calculate MCC
    data.frame(
      method = "EK03",
      event = "saccade",
      smoothing = s,
      lambda = l,
      min_fix_dur = 50,
      MCC = mltools::mcc(
        actuals = all_d2$GT_bin,
        preds = all_d2$EK_bin
      )
    )
  })
})

# Combine results
ek_saccade_results <- dplyr::bind_rows(
  unlist(ek_saccade_results, recursive = FALSE)
)

print(ek_saccade_results)

# Plot saccade MCC
ggplot(
  ek_saccade_results,
  aes(x = lambda, y = MCC, colour = smoothing)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  labs(
    title = "EK03 Saccade MCC Across Lambda Values",
    x = "Lambda",
    y = "Saccade MCC",
    colour = "Smoothing"
  ) +
  theme_classic(base_size = 14)
```
IHMM-fixation
```{r}

# IHMM parameter
ihmm_min_fix_dur <- 50

# Run IHMM fixation detection
ihmm_fixation_results <- lapply(smooth_methods, function(s) {
  
  cat("Running IHMM fixation:", s, ihmm_min_fix_dur, "\n")
  
  all_d2 <- list()
  
  for(i in 1:nrow(trial_list)) {
    
    # Select one trial
    df0 <- subset(
      dat,
      sub == trial_list$sub[i] &
        Trial_Id == trial_list$Trial_Id[i]
    )
    
    # Apply selected smoothing method
    d <- get_xy(df0, s)
    
    # Skip very short trials
    if(nrow(d) < 5) next
    
    # Run IHMM fixation detection
    algo_IHMM <- tryCatch(
      {
        eyemovements::IHMM(
          data = d,
          min_fix_dur = ihmm_min_fix_dur,
          return_saccades = FALSE,
          dva_x = 0.01865554,
          dva_y = 0.01919689
        )
      },
      error = function(e) {
        message("IHMM fixation error: ", e$message)
        return(NULL)
      }
    )
    
    # Convert detected fixations to sample-level labels
    if(is.null(algo_IHMM) || nrow(algo_IHMM) == 0) {
      
      d_IHMM <- d %>%
        dplyr::mutate(algo_state = "none")
      
    } else {
      
      d_IHMM <- d %>%
        dplyr::left_join(
          algo_IHMM %>%
            dplyr::transmute(
              fix_start,
              fix_end,
              algo_state = "fixation"
            ),
          dplyr::join_by(
            between(time, fix_start, fix_end)
          )
        ) %>%
        dplyr::mutate(
          algo_state = ifelse(
            is.na(algo_state),
            "none",
            algo_state
          )
        )
    }
    
    # Combine algorithm labels with ground truth
    d2 <- d
    d2$IHMM <- d_IHMM$algo_state
    
    # Convert labels to binary fixation classes
    d2 <- d2 %>%
      dplyr::mutate(
        GT_bin = ifelse(GT == "fixation", 1, 0),
        IHMM_bin = ifelse(IHMM == "fixation", 1, 0)
      )
    
    all_d2[[length(all_d2) + 1]] <- d2
  }
  
  # Combine all trials
  all_d2 <- dplyr::bind_rows(all_d2)
  
  # Calculate MCC
  data.frame(
    method = "IHMM",
    event = "fixation",
    smoothing = s,
    min_fix_dur = ihmm_min_fix_dur,
    MCC = mltools::mcc(
      actuals = all_d2$GT_bin,
      preds = all_d2$IHMM_bin
    )
  )
})

# Combine results
ihmm_fixation_results <- dplyr::bind_rows(ihmm_fixation_results)

print(ihmm_fixation_results)

# Plot fixation MCC across smoothing methods
ggplot(
  ihmm_fixation_results,
  aes(x = smoothing, y = MCC)
) +
  geom_col(width = 0.6) +
  geom_point(size = 2) +
  labs(
    title = "IHMM Fixation MCC Across Smoothing Methods",
    x = "Smoothing method",
    y = "Fixation MCC"
  ) +
  theme_classic(base_size = 14)
```
IHMM-saccade
```{r}
# Run IHMM saccade detection
ihmm_saccade_results <- lapply(smooth_methods, function(s) {
  
  cat("Running IHMM saccade:", s, ihmm_min_fix_dur, "\n")
  
  all_d2 <- list()
  
  for(i in 1:nrow(trial_list)) {
    
    # Select one trial
    df0 <- subset(
      dat,
      sub == trial_list$sub[i] &
        Trial_Id == trial_list$Trial_Id[i]
    )
    
    # Apply selected smoothing method
    d <- get_xy(df0, s)
    
    # Skip very short trials
    if(nrow(d) < 5) next
    
    # Run IHMM saccade detection
    algo_IHMM <- tryCatch(
      {
        eyemovements::IHMM(
          data = d,
          min_fix_dur = ihmm_min_fix_dur,
          return_saccades = TRUE,
          dva_x = 0.01865554,
          dva_y = 0.01919689
        )
      },
      error = function(e) {
        message("IHMM saccade error: ", e$message)
        return(NULL)
      }
    )
    
    # Convert detected saccades to sample-level labels
    if(is.null(algo_IHMM) || nrow(algo_IHMM) == 0) {
      
      d_IHMM <- d %>%
        dplyr::mutate(algo_state = "none")
      
    } else {
      
      d_IHMM <- d %>%
        dplyr::left_join(
          algo_IHMM %>%
            dplyr::transmute(
              sacc_start,
              sacc_end,
              algo_state = "saccade"
            ),
          dplyr::join_by(
            between(time, sacc_start, sacc_end)
          )
        ) %>%
        dplyr::mutate(
          algo_state = ifelse(
            is.na(algo_state),
            "none",
            algo_state
          )
        )
    }
    
    # Combine algorithm labels with ground truth
    d2 <- d
    d2$IHMM <- d_IHMM$algo_state
    
    # Convert labels to binary saccade classes
    d2 <- d2 %>%
      dplyr::mutate(
        GT_bin = ifelse(GT == "saccade", 1, 0),
        IHMM_bin = ifelse(IHMM == "saccade", 1, 0)
      )
    
    all_d2[[length(all_d2) + 1]] <- d2
  }
  
  # Combine all trials
  all_d2 <- dplyr::bind_rows(all_d2)
  
  # Calculate MCC
  data.frame(
    method = "IHMM",
    event = "saccade",
    smoothing = s,
    min_fix_dur = ihmm_min_fix_dur,
    MCC = mltools::mcc(
      actuals = all_d2$GT_bin,
      preds = all_d2$IHMM_bin
    )
  )
})

# Combine results
ihmm_saccade_results <- dplyr::bind_rows(ihmm_saccade_results)

print(ihmm_saccade_results)

# Plot saccade MCC across smoothing methods
ggplot(
  ihmm_saccade_results,
  aes(x = smoothing, y = MCC)
) +
  geom_col(width = 0.6) +
  geom_point(size = 2) +
  labs(
    title = "IHMM Saccade MCC Across Smoothing Methods",
    x = "Smoothing method",
    y = "Saccade MCC"
  ) +
  theme_classic(base_size = 14)
```
Run algorithms and save detected fixation events
```{r}
all_fixation_events <- list()

for(s in smooth_methods) {
  
  for(i in 1:nrow(trial_list)) {
    
    # Select one trial
    df0 <- subset(
      dat,
      sub == trial_list$sub[i] &
        Trial_Id == trial_list$Trial_Id[i]
    )
    
    # Apply selected smoothing method
    d <- get_xy(df0, s)
    
    # Skip very short trials
    if(nrow(d) < 5) next
    
    current_sub <- trial_list$sub[i]
    current_trial <- trial_list$Trial_Id[i]
    
    # IVT fixation events
    
    for(v in vel_thresholds) {
      
      algo_IVT <- tryCatch(
        {
          eyemovements::IVT(
            data = d,
            vel_threshold = v,
            min_fix_dur = min_fix_dur,
            return_saccades = FALSE,
            dva_x = 0.01865554,
            dva_y = 0.01919689
          )
        },
        error = function(e) return(NULL)
      )
      
      if(!is.null(algo_IVT) && nrow(algo_IVT) > 0) {
        
        ev <- algo_IVT %>%
          dplyr::transmute(
            method = "IVT",
            event = "fixation",
            smoothing = s,
            parameter = paste0("vel_", v),
            sub = current_sub,
            Trial_Id = current_trial,
            fix_start = fix_start,
            fix_end = fix_end,
            duration = fix_end - fix_start
          )
        
        all_fixation_events[[length(all_fixation_events) + 1]] <- ev
      }
    }
    

    # IDT fixation events
    for(disp in disp_thresholds) {
      
      algo_IDT <- tryCatch(
        {
          eyemovements::IDT(
            data = d,
            disp_threshold = disp,
            window_threshold = window_threshold,
            return_saccades = FALSE,
            dva_x = 0.01865554,
            dva_y = 0.01919689
          )
        },
        error = function(e) return(NULL)
      )
      
      if(!is.null(algo_IDT) && nrow(algo_IDT) > 0) {
        
        ev <- algo_IDT %>%
          dplyr::transmute(
            method = "IDT",
            event = "fixation",
            smoothing = s,
            parameter = paste0("disp_", disp),
            sub = current_sub,
            Trial_Id = current_trial,
            fix_start = fix_start,
            fix_end = fix_end,
            duration = fix_end - fix_start
          )
        
        all_fixation_events[[length(all_fixation_events) + 1]] <- ev
      }
    }
    
  
    # EK03 fixation events
    
    for(l in lambdas) {
      
      algo_EK <- tryCatch(
        {
          eyemovements::EngbertKliegl03(
            data = d,
            lambda = l,
            min_fix_dur = 50,
            return_saccades = FALSE,
            dva_x = 0.01865554,
            dva_y = 0.01919689
          )
        },
        error = function(e) return(NULL)
      )
      
      if(!is.null(algo_EK) && nrow(algo_EK) > 0) {
        
        ev <- algo_EK %>%
          dplyr::transmute(
            method = "EK03",
            event = "fixation",
            smoothing = s,
            parameter = paste0("lambda_", l),
            sub = current_sub,
            Trial_Id = current_trial,
            fix_start = fix_start,
            fix_end = fix_end,
            duration = fix_end - fix_start
          )
        
        all_fixation_events[[length(all_fixation_events) + 1]] <- ev
      }
    }
    
   
    # IHMM fixation events
    algo_IHMM <- tryCatch(
      {
        eyemovements::IHMM(
          data = d,
          min_fix_dur = 50,
          return_saccades = FALSE,
          dva_x = 0.01865554,
          dva_y = 0.01919689
        )
      },
      error = function(e) return(NULL)
    )
    
    if(!is.null(algo_IHMM) && nrow(algo_IHMM) > 0) {
      
      ev <- algo_IHMM %>%
        dplyr::transmute(
          method = "IHMM",
          event = "fixation",
          smoothing = s,
          parameter = "none",
          sub = current_sub,
          Trial_Id = current_trial,
          fix_start = fix_start,
          fix_end = fix_end,
          duration = fix_end - fix_start
        )
      
      all_fixation_events[[length(all_fixation_events) + 1]] <- ev
    }
  }
}

all_fixation_events <- dplyr::bind_rows(all_fixation_events)

print(all_fixation_events)
```
Run algorithms and save detected saccade events
```{r}

all_saccade_events <- list()

for(s in smooth_methods) {
  
  for(i in 1:nrow(trial_list)) {
    
    # Select one trial
    df0 <- subset(
      dat,
      sub == trial_list$sub[i] &
        Trial_Id == trial_list$Trial_Id[i]
    )
    
    # Apply selected smoothing method
    d <- get_xy(df0, s)
    
    # Skip very short trials
    if(nrow(d) < 5) next
    
    current_sub <- trial_list$sub[i]
    current_trial <- trial_list$Trial_Id[i]
    

    # IVT saccade events
    for(v in vel_thresholds) {
      
      algo_IVT <- tryCatch(
        {
          eyemovements::IVT(
            data = d,
            vel_threshold = v,
            min_fix_dur = min_fix_dur,
            return_saccades = TRUE,
            dva_x = 0.01865554,
            dva_y = 0.01919689
          )
        },
        error = function(e) return(NULL)
      )
      
      if(!is.null(algo_IVT) && nrow(algo_IVT) > 0) {
        
        ev <- algo_IVT %>%
          dplyr::transmute(
            method = "IVT",
            event = "saccade",
            smoothing = s,
            parameter = paste0("vel_", v),
            sub = current_sub,
            Trial_Id = current_trial,
            sacc_start = sacc_start,
            sacc_end = sacc_end,
            duration = sacc_end - sacc_start
          )
        
        all_saccade_events[[length(all_saccade_events) + 1]] <- ev
      }
    }
    

    # IDT saccade events
    for(disp in disp_thresholds) {
      
      algo_IDT <- tryCatch(
        {
          eyemovements::IDT(
            data = d,
            disp_threshold = disp,
            window_threshold = window_threshold,
            return_saccades = TRUE,
            dva_x = 0.01865554,
            dva_y = 0.01919689
          )
        },
        error = function(e) return(NULL)
      )
      
      if(!is.null(algo_IDT) && nrow(algo_IDT) > 0) {
        
        ev <- algo_IDT %>%
          dplyr::transmute(
            method = "IDT",
            event = "saccade",
            smoothing = s,
            parameter = paste0("disp_", disp),
            sub = current_sub,
            Trial_Id = current_trial,
            sacc_start = sacc_start,
            sacc_end = sacc_end,
            duration = sacc_end - sacc_start
          )
        
        all_saccade_events[[length(all_saccade_events) + 1]] <- ev
      }
    }
    
   
    # EK03 saccade events
    for(l in lambdas) {
      
      algo_EK <- tryCatch(
        {
          eyemovements::EngbertKliegl03(
            data = d,
            lambda = l,
            min_fix_dur = 50,
            return_saccades = TRUE,
            dva_x = 0.01865554,
            dva_y = 0.01919689
          )
        },
        error = function(e) return(NULL)
      )
      
      if(!is.null(algo_EK) && nrow(algo_EK) > 0) {
        
        ev <- algo_EK %>%
          dplyr::transmute(
            method = "EK03",
            event = "saccade",
            smoothing = s,
            parameter = paste0("lambda_", l),
            sub = current_sub,
            Trial_Id = current_trial,
            sacc_start = sacc_start,
            sacc_end = sacc_end,
            duration = sacc_end - sacc_start
          )
        
        all_saccade_events[[length(all_saccade_events) + 1]] <- ev
      }
    }
    
  
    # IHMM saccade events
    algo_IHMM <- tryCatch(
      {
        eyemovements::IHMM(
          data = d,
          min_fix_dur = 50,
          return_saccades = TRUE,
          dva_x = 0.01865554,
          dva_y = 0.01919689
        )
      },
      error = function(e) return(NULL)
    )
    
    if(!is.null(algo_IHMM) && nrow(algo_IHMM) > 0) {
      
      ev <- algo_IHMM %>%
        dplyr::transmute(
          method = "IHMM",
          event = "saccade",
          smoothing = s,
          parameter = "none",
          sub = current_sub,
          Trial_Id = current_trial,
          sacc_start = sacc_start,
          sacc_end = sacc_end,
          duration = sacc_end - sacc_start
        )
      
      all_saccade_events[[length(all_saccade_events) + 1]] <- ev
    }
  }
}

all_saccade_events <- dplyr::bind_rows(all_saccade_events)

print(all_saccade_events)

```



MCC Results
```{r}

# Combine IVT fixation + saccade MCC results
ivt_mcc <- bind_rows(
  ivt_fixation_results,
  ivt_saccade_results
) %>%
  mutate(
    parameter_type = "vel_threshold",
    parameter_value = vel_threshold
  ) %>%
  select(
    method, event, smoothing,
    parameter_type, parameter_value,
    MCC
  )

# Combine IDT fixation + saccade MCC results
idt_mcc <- bind_rows(
  idt_fixation_results,
  idt_saccade_results
) %>%
  mutate(
    parameter_type = "disp_threshold",
    parameter_value = disp_threshold
  ) %>%
  select(
    method, event, smoothing,
    parameter_type, parameter_value,
    MCC
  )

# Combine EK03 fixation + saccade MCC results
ek_mcc <- bind_rows(
  ek_fixation_results,
  ek_saccade_results
) %>%
  mutate(
    parameter_type = "lambda",
    parameter_value = lambda
  ) %>%
  select(
    method, event, smoothing,
    parameter_type, parameter_value,
    MCC
  )

# Combine IHMM fixation + saccade MCC results
ihmm_mcc <- bind_rows(
  ihmm_fixation_results,
  ihmm_saccade_results
) %>%
  mutate(
    parameter_type = "none",
    parameter_value = NA_real_
  ) %>%
  select(
    method, event, smoothing,
    parameter_type, parameter_value,
    MCC
  )

# Combine all algorithm MCC results
all_mcc_results <- bind_rows(
  ivt_mcc,
  idt_mcc,
  ek_mcc,
  ihmm_mcc
)


# Save to CSV
write_csv(
  all_mcc_results,
  "results/all_algorithm_mcc_results.csv"
)
```

### Event-level agreement (Wasserstein distance)

To quantify how well each algorithm's event-level characteristics match the human-labelled ground truth, each trial was summarised by three features: the number of events, the mean event duration, and the SD of event duration. The multivariate Wasserstein distance between the algorithm's and the ground truth's per-trial distributions was then computed, after standardising all three features using the ground-truth mean and SD. The distance was transformed to a similarity score $W_{\mathrm{multi}} = 1 / (1 + d)$, where a value of 1 indicates identical distributions.

```{r wasserstein}
# Event-level Wasserstein distance 

compute_wasserstein_full <- function(gt_mat, algo_mat, p = 1) {
  valid <- complete.cases(gt_mat) & complete.cases(algo_mat)
  gt_mat   <- gt_mat[valid, , drop = FALSE]
  algo_mat <- algo_mat[valid, , drop = FALSE]
  if (nrow(gt_mat) < 3 || nrow(algo_mat) < 3)
    return(list(W_multi = NA, W_n = NA, W_mean = NA, W_sd = NA))

  gt_center <- colMeans(gt_mat)
  gt_scale  <- apply(gt_mat, 2, sd)
  gt_scale[gt_scale == 0] <- 1

  gt_scaled   <- scale(gt_mat,   center = gt_center, scale = gt_scale)
  algo_scaled <- scale(algo_mat, center = gt_center, scale = gt_scale)

  w_dist <- function(g, a, cols) {
    d <- T4transport::wasserstein(
      scale(g[, cols, drop = FALSE], center = gt_center[cols], scale = gt_scale[cols]),
      scale(a[, cols, drop = FALSE], center = gt_center[cols], scale = gt_scale[cols]),
      p = p
    )$distance
    1 / (1 + d)
  }

  list(
    W_multi = 1 / (1 + T4transport::wasserstein(gt_scaled, algo_scaled, p = p)$distance),
    W_n     = w_dist(gt_mat, algo_mat, 1),   # event count
    W_mean  = w_dist(gt_mat, algo_mat, 2),   # mean duration
    W_sd    = w_dist(gt_mat, algo_mat, 3)    # SD duration
  )
}

#Traditional algorithms: per (method, smoothing, parameter) 
run_wasserstein_trad <- function(comp_df, event_type) {
  if (event_type == "fixation") {
    n_c  <- "algo_n_fixations";  m_c  <- "algo_mean_fix_dur";  s_c  <- "algo_sd_fix_dur"
    gn_c <- "GT_n_fixations";    gm_c <- "GT_mean_fix_dur";    gs_c <- "GT_sd_fix_dur"
  } else {
    n_c  <- "algo_n_saccades";   m_c  <- "algo_mean_sacc_dur"; s_c  <- "algo_sd_sacc_dur"
    gn_c <- "GT_n_saccades";     gm_c <- "GT_mean_sacc_dur";   gs_c <- "GT_sd_sacc_dur"
  }

  comp_df %>%
    group_by(method, smoothing, parameter) %>%
    group_modify(~ {
      gt   <- .x %>% select(all_of(c(gn_c, gm_c, gs_c))) %>% as.matrix()
      algo <- .x %>% select(all_of(c(n_c,  m_c,  s_c)))  %>% as.matrix()
      w <- compute_wasserstein_full(gt, algo)
      data.frame(W_multi = w$W_multi, W_n = w$W_n, W_mean = w$W_mean, W_sd = w$W_sd)
    }) %>%
    ungroup() %>%
    mutate(event = event_type)
}

fix_comp <- read_csv(
  here("manuscript", "results", "event_level_rmse", "fixation_event_level_comparison.csv"),
  show_col_types = FALSE
)
sac_comp <- read_csv(
  here("manuscript", "results", "event_level_rmse", "saccade_event_level_comparison.csv"),
  show_col_types = FALSE
)

W_fix <- run_wasserstein_trad(fix_comp, "fixation")
W_sac <- run_wasserstein_trad(sac_comp, "saccade")

# CNN-BLSTM per-trial event-level data (all smoothing types)
ml_trial_raw    <- read_csv(here("..", "results", "ml_event_level_by_trial_raw.csv"),       show_col_types = FALSE) %>% mutate(smoothing = "raw")
ml_trial_mean   <- read_csv(here("..", "results", "ml_event_level_by_trial_mean.csv"),      show_col_types = FALSE) %>% mutate(smoothing = "mean")
ml_trial_median <- read_csv(here("..", "results", "ml_event_level_by_trial_median.csv"),    show_col_types = FALSE) %>% mutate(smoothing = "median")
ml_trial_sg     <- read_csv(here("..", "results", "ml_event_level_by_trial_sg.csv"),        show_col_types = FALSE) %>% mutate(smoothing = "sg_p3_n7")
ml_trial_sg5    <- read_csv(here("..", "results", "ml_event_level_by_trial_sg_p5_n23.csv"), show_col_types = FALSE) %>% mutate(smoothing = "sg_p5_n23")

ml_trial <- bind_rows(ml_trial_raw, ml_trial_mean, ml_trial_median, ml_trial_sg, ml_trial_sg5) %>%
  mutate(method = "CNN-BLSTM", parameter = "none")

ml_wasserstein <- ml_trial %>%
  group_by(method, smoothing, parameter, event_type) %>%
  group_modify(~ {
    gt   <- .x %>% select(gt_event_count, gt_mean_duration_ms, gt_sd_duration_ms) %>% as.matrix()
    algo <- .x %>% select(pred_event_count, pred_mean_duration_ms, pred_sd_duration_ms) %>% as.matrix()
    w <- compute_wasserstein_full(gt, algo)
    data.frame(W_multi = w$W_multi, W_n = w$W_n, W_mean = w$W_mean, W_sd = w$W_sd)
  }) %>%
  ungroup() %>%
  rename(event = event_type)

W_all <- bind_rows(W_fix, W_sac, ml_wasserstein)

#Best configuration per method x event 
table_s2 <- W_all %>%
  filter(!is.na(W_multi)) %>%
  group_by(method, event) %>%
  slice_max(W_multi, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(method, event, smoothing, parameter,
         `W_multi (overall)` = W_multi,
         `W_n (count)`       = W_n,
         `W_mean (duration)` = W_mean,
         `W_sd (SD)`         = W_sd) %>%
  arrange(event, desc(`W_multi (overall)`))

table_s2
```


## Results


```{r}

# Plot fixation detections for one trial
# GT + IVT + IDT + EK + IHMM


selected_sub <- 11
selected_trial <- 114

d0 <- subset(
  dat,
  sub == selected_sub &
    Trial_Id == selected_trial
)

# Webcam data for algorithms
d <- d0 %>%
  transmute(
    x = x,
    y = y,
    time = time_start,
    GT = ground_truth
  )

# EyeLink data only for trace plot
d_el <- d0 %>%
  transmute(
    x = el_x,
    y = el_y,
    time = time_start,
    GT = ground_truth
  )

# Smooth both traces
d <- SmoothSamples(d, method = "Mean", window_size = 3)
d_el <- SmoothSamples(d_el, method = "Mean", window_size = 3)


# IVT fixation detection
algo_IVT <- IVT(
  data = d,
  vel_threshold = 50,
  min_fix_dur = 50,
  return_saccades = FALSE,
  dva_x = 0.01865554,
  dva_y = 0.01919689
)

if(nrow(algo_IVT) == 0) {
  
  d_IVT <- d %>%
    mutate(algo_state = "none") %>%
    select(x, y, time, GT, algo_state)
  
} else {
  
  d_IVT <- d %>%
    left_join(
      algo_IVT %>%
        transmute(
          fix_start,
          fix_end,
          algo_state = "fixation"
        ),
      join_by(
        between(time, fix_start, fix_end)
      )
    ) %>%
    mutate(
      algo_state = ifelse(is.na(algo_state), "none", algo_state)
    ) %>%
    select(x, y, time, GT, algo_state)
}


# IDT fixation detection
algo_IDT <- IDT(
  data = d,
  disp_threshold = 1,
  window_threshold = 50,
  dva_x = 0.01865554,
  dva_y = 0.01919689
)

if(nrow(algo_IDT) == 0) {
  
  d_IDT <- d %>%
    mutate(algo_state = "none") %>%
    select(x, y, time, GT, algo_state)
  
} else {
  
  d_IDT <- d %>%
    left_join(
      algo_IDT %>%
        transmute(
          fix_start,
          fix_end,
          algo_state = "fixation"
        ),
      join_by(
        between(time, fix_start, fix_end)
      )
    ) %>%
    mutate(
      algo_state = ifelse(is.na(algo_state), "none", algo_state)
    ) %>%
    select(x, y, time, GT, algo_state)
}


# EK03 fixation detection
algo_EK <- EngbertKliegl03(
  data = d,
  lambda = 3,
  min_fix_dur = 50,
  return_saccades = FALSE,
  dva_x = 0.01865554,
  dva_y = 0.01919689
)

if(nrow(algo_EK) == 0) {
  
  d_EK <- d %>%
    mutate(algo_state = "none") %>%
    select(x, y, time, GT, algo_state)
  
} else {
  
  d_EK <- d %>%
    left_join(
      algo_EK %>%
        transmute(
          fix_start,
          fix_end,
          algo_state = "fixation"
        ),
      join_by(
        between(time, fix_start, fix_end)
      )
    ) %>%
    mutate(
      algo_state = ifelse(is.na(algo_state), "none", algo_state)
    ) %>%
    select(x, y, time, GT, algo_state)
}


# IHMM fixation detection
algo_IHMM <- IHMM(
  data = d,
  min_fix_dur = 50,
  return_saccades = FALSE,
  dva_x = 0.01865554,
  dva_y = 0.01919689
)

if(nrow(algo_IHMM) == 0) {
  
  d_IHMM <- d %>%
    mutate(algo_state = "none") %>%
    select(x, y, time, GT, algo_state)
  
} else {
  
  d_IHMM <- d %>%
    left_join(
      algo_IHMM %>%
        transmute(
          fix_start,
          fix_end,
          algo_state = "fixation"
        ),
      join_by(
        between(time, fix_start, fix_end)
      )
    ) %>%
    mutate(
      algo_state = ifelse(is.na(algo_state), "none", algo_state)
    ) %>%
    select(x, y, time, GT, algo_state)
}


# Combine detected states
d2 <- d
d2$IVT <- d_IVT$algo_state
d2$IDT <- d_IDT$algo_state
d2$EK <- d_EK$algo_state
d2$IHMM <- d_IHMM$algo_state


# Prepare trace data
trace_dat_w <- d %>%
  select(time, x, y) %>%
  pivot_longer(
    cols = c(x, y),
    names_to = "signal",
    values_to = "position"
  ) %>%
  mutate(tracker = "Webcam")

trace_dat_el <- d_el %>%
  select(time, x, y) %>%
  pivot_longer(
    cols = c(x, y),
    names_to = "signal",
    values_to = "position"
  ) %>%
  mutate(tracker = "EyeLink")

trace_dat <- bind_rows(trace_dat_w, trace_dat_el)

P_trace <- ggplot(
  trace_dat,
  aes(
    x = time,
    y = position,
    colour = tracker,
    linetype = signal
  )
) +
  geom_line(linewidth = 0.5, alpha = 0.8) +
  labs(
    x = NULL,
    y = "Position (px)",
    colour = "Tracker",
    linetype = "Signal"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "top",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )


# Prepare event data
state_dat <- d2 %>%
  select(time, GT, IVT, IDT, EK, IHMM) %>%
  mutate(
    GT = ifelse(GT == "fixation", "fixation", "none")
  ) %>%
  pivot_longer(
    cols = c(GT, IVT, IDT, EK, IHMM),
    names_to = "source",
    values_to = "state"
  ) %>%
  mutate(
    source = factor(
      source,
      levels = c("IHMM", "EK", "IDT", "IVT", "GT")
    )
  )


# Convert continuous fixation states into events
event_dat <- state_dat %>%
  arrange(source, time) %>%
  group_by(source) %>%
  mutate(
    event = data.table::rleid(state)
  ) %>%
  group_by(source, event, state) %>%
  summarise(
    start = min(time, na.rm = TRUE),
    end = max(time, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::filter(state == "fixation") %>%
  mutate(
    source_num = as.numeric(source),
    ymin = source_num - 0.35,
    ymax = source_num + 0.35
  )

P_events <- ggplot(event_dat) +
  geom_rect(
    aes(
      xmin = start,
      xmax = end,
      ymin = ymin,
      ymax = ymax,
      fill = source
    ),
    colour = "black",
    linewidth = 0.25,
    alpha = 0.85
  ) +
  scale_y_continuous(
    breaks = seq_along(levels(event_dat$source)),
    labels = levels(event_dat$source),
    limits = c(0.5, length(levels(event_dat$source)) + 0.5)
  ) +
  labs(
    x = "Trial time",
    y = NULL,
    fill = NULL
  ) +
  theme_classic(base_size = 14) +
  theme(
    axis.text.y = element_text(face = "bold"),
    legend.position = "none"
  )

# Combine trace and event plots
P_trace / P_events +
  plot_layout(
    heights = c(3, 1.5)
  )
```

