---
title: "Manuscript_draft"
output: html_document
date: "2026-04-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(here)
library(tidyverse)
```

## Introduction

## Method

### Dataset

```{r, include= F }

dat <- read_csv(here("data/manual_labels", "webdata_manual_labels.csv"))

n_before_blinks<- nrow(dat)

# Clean data:
dat<- subset(dat,conf>0 & el_pupil>0)# remove blinks:

n_after_blinks<- nrow(dat)

blink_loss<- n_after_blinks/ n_before_blinks

labels<- read_csv(here("data", "manual_ground_truth.csv"))

n_fixations<- length(which(labels$label=="fixation"))
n_saccades<- length(which(labels$label=="saccade"))
n_pso<- length(which(labels$label=="pso"))
n_unclear<- length(which(labels$label=="unclear"))

```

The data contained `r n_before_blinks` raw sample observatations. After blinks were removed, `r n_after_blinks` remained (`r round(blink_loss*100,2)` %). The data manually labelled ground truth contained `r n_fixations` fixations, `r n_saccades` saccades, `r n_pso` PSOs, and `r n_unclear` unclear events (noise). 


```{r}
labels %>% 
  filter(label!= 'pso'& label!= 'unclear')%>%
  group_by(label)%>%
  ggplot(aes(x= label, y= duration, fill= label))+
  geom_violin(alpha= 0.4)+
  geom_boxplot(width= 0.1)+
  #facet_wrap(~label, scales = "free_y")+
  theme_bw(18)
```


### Saccade Detection algorithms

#### I-VT
#### I-DT
#### Engbert & Kliegl, 2003 [EK03]
#### Machine learning algorithm
#### Others...


### Sample Smoothing methods

Raw, Mean filter, Median filter, Savitsky-Golay, Others? 


## Results


